Intervention Bandits

### 2014-11-06 causal-bandit/c-bandit.tex
started to add proof
fixed some typos in proof
updates c-bandit to include experimental figures

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}

@@

\title{Causal Inference and Bandits}
\author{Blah blah}

@@

We are considering a variant of the multi-armed bandit problem.
Let $p \in [0,1]^K$ be a fixed and known vector.
In each time-step $t$ 
\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

@@

Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
\eq{
n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
n_a        &= \begin{cases} 
  n_{i,j}                                           & \text{if } a = i \\
  \min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
\end{cases}
}
\begin{theorem} With probability at least $1 - \delta$ we have that:
$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
where $\beta > 0$ is some constant.
\end{theorem}

@@

\section{Theorems}
\section{Experiments}
\section{Conclusion}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{times}
\usepackage{fullpage}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{accents}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage[ruled]{algorithm}
\usepackage{algpseudocode}
\usepackage{dsfont}
\usepackage[bf]{caption}
\usepackage{hyperref}
\hypersetup{
    bookmarks=true,         % show bookmarks bar?
    unicode=false,          % non-Latin characters in AcrobatÕs bookmarks
    pdftoolbar=true,        % show AcrobatÕs toolbar?
    pdfmenubar=true,        % show AcrobatÕs menu?
    pdffitwindow=false,     % window fit to page when opened
    pdfstartview={FitH},    % fits the width of the page to the window
    pdftitle={My title},    % title
    pdfauthor={Author},     % author
    pdfsubject={Subject},   % subject of the document
    pdfcreator={Creator},   % creator of the document
    pdfproducer={Producer}, % producer of the document
    pdfkeywords={keyword1} {key2} {key3}, % list of keywords
    pdfnewwindow=true,      % links in new window
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage[capitalize]{cleveref}
\usepackage{graphicx}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MACROS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\defined}{\vcentcolon =}
\newcommand{\rdefined}{=\vcentcolon}
\newcommand{\E}{\mathbb E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\calF}{\mathcal F}
\newcommand{\sr}[1]{\stackrel{#1}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\ind}[1]{\mathds{1}\!\!\set{#1}}
\newcommand{\argmax}{\operatornamewithlimits{arg\,max}}
\newcommand{\argmin}{\operatornamewithlimits{arg\,min}}
\newcommand{\floor}[1]{\left \lfloor {#1} \right\rfloor}
\newcommand{\ceil}[1]{\left \lceil {#1} \right\rceil}
\newcommand{\eqn}[1]{\begin{align}#1\end{align}}
\newcommand{\eq}[1]{\begin{align*}#1\end{align*}}
\newcommand{\Ber}{\operatorname{Bernoulli}}
\renewcommand{\P}[1]{\operatorname{P}\left\{#1\right\}}

@@

\title{Intervention Bandits}
\author{Blah blah}

@@

Assume we have a known causal model with binary variables $\boldsymbol{X} = \{X_{1}..X_{K}\}$ that independently cause a target variable of interest $Y$. We can run sequential experiments on the system, where at each timestep $t$ we can select a variable on which to intervene and then we observe the complete result, $(\boldsymbol{X}_{t},Y_{t})$ This problem can be viewed as a variant of the multi-armed bandit problem.

Let $p \in [0,1]^K$ be a fixed and known vector. In each time-step $t$:

\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|X_i = j]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

@@

Fix some time-step $t$ and $i \in \set{1,\ldots,K}$ and $j \in \set{0,1}$.
We define estimator $\hat \mu_t$ of $\mu_{i,j}$ by
\eq{
m_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b, Y_t = 1} \\
n_{a,b}    &= \sum_{s=1}^t \ind{I_t = i, J_t = j, X_a = b} \\
\hat \mu_t &= \eta_i \frac{m_{i,j}}{n_{i,j}}  + \sum_{a \neq i} \eta_a \left[p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}}\right] \\
\eta_a     &= \frac{n_a}{\sum_{a=1}^K n_a} \\
n_a        &= \begin{cases} 
  n_{i,j}                                           & \text{if } a = i \\
  \frac{1}{2}\min\set{\frac{n_{a,1}}{p_a}, \frac{n_{a,0}}{1-p_a}}  & \text{otherwise}
\end{cases}
}
\begin{theorem} With probability at least $1 - \delta$ we have that:
$\displaystyle \left|\hat \mu_t - \mu\right| \leq \sqrt{\frac{\beta}{\sum_{a} n_a} \log{1 \over \delta}}$,
where $\beta > 0$ is some constant.
\end{theorem}

\begin{proof}
First note that $n_{a,b}$ is a random variable that is bounded by $t$ for all $a,b$.
We use the short-hand $\mu^{a,b}_{i,j} = \E[q(X) | X_i = j, X_a = b]$. Then
\eq{
\mu_{i,j} = p_a \mu_{i,j}^{a,1} + (1 - p_a) \mu_{i,j}^{a,0}\,.
}
Now we can apply Hoeffding's bound and the union bound to show that
\eq{
\P{\left|\frac{m_{a,b}}{n_{a,b}} - \mu_{i,j}^{a,b}\right| \geq \sqrt{\frac{1}{2n_{a,b}} \log \frac{4t}{\delta}}} \leq \frac{\delta}{2}\,.
}
Therefore by the union bound
\eq{
\P{\left|p_a \frac{m_{a,1}}{n_{a,1}} + (1 - p_a) \frac{m_{a,0}}{n_{a,0}} - \mu_{i,j}\right| \geq p_a \sqrt{\frac{1}{2n_{a,1}} \log\frac{4t}{\delta}}
+ (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}} \leq \delta\,
}
Now by Jensen's inequality 
\eq{
p_a \sqrt{\frac{1}{2n_{a,1}}\log\frac{4t}{\delta}} + (1 - p_a) \sqrt{\frac{1}{2n_{a,0}}\log\frac{4t}{\delta}}
&\leq \sqrt{\left(\frac{p_a}{2n_{a,1}} + \frac{1-p_a}{2n_{a,0}}\right)\log \frac{4t}{\delta}} \\
&\leq \sqrt{\max\set{\frac{p_a}{n_{a,1}}, \frac{1-p_a}{n_{a,0}}} \log\frac{4t}{\delta}} \\
&= \sqrt{{1 \over 2n_a} \log\frac{4t}{\delta}}\,.
}
Similarly,
\eq{
\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{4t}{\delta}}} \leq
\P{\left|\frac{m_{i,j}}{n_{i,j}} - \mu_{i,j}\right| \geq \sqrt{\frac{1}{2n_a} \log\frac{2t}{\delta}}} \leq \delta\,.
}
\end{proof}

@@

\section{Theorems}
\section{Experiments}

\begin{figure}[h]
\caption{Testing values of beta - 100 experiments for each value of beta, with a 4 armed bandit.}
\includegraphics[width=0.7\textwidth]{regretvsbeta.pdf}
\end{figure}

Simululations to compare the performance of standard UCB with our modified algorithm. For each number of arms, $100$ bandits of each type were created and run upto to a horizon of $1000$ timesteps. The mean regret and its standard error from these simulations is plotted in figure \ref{fig:regretvsarms} The true data was generated from a model where:

\eq{
&p = [0.5]^{K}\\
&q(\boldsymbol{X}) = \begin{cases}0.5 & \text{if $X_{1}=0$}\\ 0.6 & \text{otherwise} \end{cases}
}

For small number of arms there is not much difference. As the number of arms increases, cusb shows significantly lower regret. As the number of arms increases above 100 the regret of the two algorithms begins to converge again. I suspect this is due to the fixed time horizon of $1000$ steps. With a time horizon of $100$ steps the two algorithms begin to converge again at a smaller number of arms.

\begin{figure}[h]
\caption{Shows relationship between the regret and the number of bandit arms $2K$ for the standard UCB algorithm and our modified algorithm.  }
\label{fig:regretvsarms}
\includegraphics[width=0.9\textwidth]{regretvsarms.pdf}
\end{figure}

\section{Conclusion}
